Physcraper: Body

### 2020-07-27 2-alt-1-intro.Rmd
intro clean 4

>>>>

Physcraper is a Python encoded pipeline that focuses on database interoperability and provides programmatic access to the GenBank, TreeBASE and OpenTree public databases, to automatize taxonomic name matching across virtually any  biological database, so users can perform any kind of downstream analysis straightforwardly. It's main goal is to link molecular data to alignments and phylogenies, as well as other biological databases to place new biological data in an evolutionary context.

### 2020-07-27 2-body.Rmd
intro clean 4

The prevalence of taxonomic idiosincracies across databases represent a big challenge to automatically connect data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified taxonomy is necessary. The main aim of the Open Tree of Life (OpenTree from now on) project
<!-- [project](https://opentreeoflife.github.io/)  -->
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's "synthetic" tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies/integrates taxonomy from various databases [@rees2017automated] such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank [@benson2000genbank; @wheeler2000database], the Global biodiversity Information Facility [GBIF; @secretariat2017gbif], the Encyclopedia of Life [EOL; @wilson2003encyclopedia], etc., providing a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses. Species tree reconstructions from multi-gene data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships [@song2012resolving].
<!-- ROJAS ET AL. bats paper, take citations from there -->
While genomics have allowed us to obtain large quantities of molecular data from one organism, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing (RADSeq), single nucleotide polymorphisms (SNPs), microsatellites and ultra-conserved elements (UCEs).
Having a variety of approaches has resulted in largely non-overlapping genomic data sets across taxa. Moreover, some types of genomic data are not suitable for phylogenetic inference at any scale. For example, phylogenetic distance can cause allelic dropout at deeper divergences with RADSeq data sets [@eaton2016misconceptions].
<!-- allelic dropout -->
In contrast, decades of single locus sequencing have generated homologous DNA data
<!-- across large phylogenetic scales  -->
that can be used for phylogenetic reconstruction at many scales. Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy [about `r round(236023/240708, digits=4)*100`% of taxa; @sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand" but it requires huge amounts of time and it is a mostly non-reproducible method. Computer pipelines that make DNA sequence search faster and more efficient, as well as more reproducible have been utilized to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
Still, there is a large tradition in phylogenetics of building molecular character matrices of few, thoughtfully curated markers, as manual curation of alignments often improves phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accesible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), as well as an untapped source of valuable expert knowledge with the potential to continuously enrich and update phylogenetic relationships in several different regions of the tree of life, by incorporating data from molecular databases that has not been included in any public phylogenetic estimate.

The Phylesystem (OpenTree's datastore) [@mctavish2015phylesystem] has integrated all phylogenies from TreeBASE, while saving metadata linking trees to their corresponding alignment repository in TreeBASE. If there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, this metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link.

Physcraper provides programmatic access to the GenBank, TreeBASE and OpenTree Phylesystem public databases, to link molecular data to alignments and phylogenies, as well as other biological databases to place new biological data in an evolutionary context.

>>>>

The prevalence of taxonomic idiosincracies across databases represent a big challenge to automatically connect data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified taxonomy is needed. The main aim of the Open Tree of Life (OpenTree from now on) project
<!-- [project](https://opentreeoflife.github.io/)  -->
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's "synthetic" tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases [@rees2017automated] such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank [@benson2000genbank; @wheeler2000database], the Global Biodiversity Information Facility [GBIF; @secretariat2017gbif], and the Encyclopedia of Life [EOL; @wilson2003encyclopedia], providing a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses. Species tree reconstructions from multi-gene data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships [@song2012resolving].
<!-- ROJAS ET AL. bats paper, take citations from there -->
While genomics have allowed us to obtain large quantities of molecular data from one organism, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing (RADSeq), single nucleotide polymorphisms (SNPs), microsatellites, and ultra-conserved elements (UCEs).
Having a variety of approaches has resulted in largely non-overlapping genomic data sets across taxa. Moreover, some types of genomic data are not suitable for phylogenetic inference at any scale. For example, phylogenetic distance can cause allelic dropout at deeper divergences with RADSeq data sets [@eaton2016misconceptions].
<!-- allelic dropout -->
In contrast, decades of single locus sequencing have generated homologous DNA data
<!-- across large phylogenetic scales  -->
that can be used for phylogenetic reconstruction at many scales. Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy at the time [about `r round(236023/240708, digits=4)*100`% of taxa; @sanderson2008phylota]. Assembling a DNA alignment from such a massive database can be done "by hand", but it requires huge amounts of time and it is mostly a non-reproducible method. Computer pipelines that make DNA sequence search faster and more efficient, as well as more reproducible have been utilized to study evolutionary relationships among a variety of organisms [e.g., @smith2009mega; @antonelli2017toward; @izquierdo2014pumper].
<!-- *maybe a box of all pipelines?* -->
Still, there is a large tradition in phylogenetics of building molecular character matrices of few, thoughtfully curated markers, as manual curation of alignments often improves phylogenetic reconstructions [@fragoso2017pilot].

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as "jump-start" alignments [@morrison2006multiple]. The TreeBASE database [@piel2009treebase] hosts about 8, 200 publicly accesible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa [(see TreeBASE's website about)](https://www.treebase.org/treebase-web/home.html#:~:text=TreeBASE%20is%20produced%20and%20governed,mapped%20to%20104%2C593%20distinct%20taxa.), representing an untapped source of valuable expert knowledge with the potential to continuously enrich and update phylogenetic relationships in several different regions of the tree of life, by incorporating data from molecular databases that has not been included in any public phylogenetic estimate. The phylogenies associated to all these alignments have been integrated to the OpenTree's datastore, the Phylesystem [@mctavish2015phylesystem], and metadata linking these trees to their corresponding alignment repository in TreeBASE is available.
<!-- However, if there are multiple alignments and trees, TreeBASE does not always indicate how they are connected. Hence, Phylesystem's metadata only provides a loose means of linking the tree with the exact alignment that generated it and a human curator must make the link. -->

We propose Physcraper as a Python encoded pipeline that relies on programmatic access protocols to link molecular data from GenBank, to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to automatically update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for custom comparison of phylogenetic hypotheses.
It focuses on database interoperability, by automatizing taxonomic name matching across biological databases, so users can perform any downstream analysis straightforwardly, as well as placing new data from virtually any biological database in an evolutionary context.
